Recurrent event analysis in the presence of real-time high frequency data via random subsampling

### 2019-07-27 write-up/fda-recurrent.tex
Latest version of write-up, fixed some typos

An estimator is design-unbiased if its expectation is equal to that
parameter under the probability distribution induced by the sampling
design~\citep{Cassel1977}. Let~$D_i \subset [0,t_i]$ denote the set of
subsampled points.  Note, by design, this set is distinct from the set
of event times, i.e.,~$\bfT_i \cap D_i = \emptyset$.  Under
subsampling via $\pi_i (t)$, one may compute a Horvitz-Thompson
estimator of the derivative of the cumulative hazard:
\[
\hat H_{i}^{(1)} (\tau_i; \theta) = \sum_{u \in D_i} \frac{
  h^{(1)}_i (u; \theta) }{ \pi_i (s) }
\]
An alternative design-unbiased estimator of the derivative of the
cumulative hazards is given by
\begin{equation}
\label{eq:WPest}
\hat H_{i}^{(1)} (\tau_i; \theta) = \sum_{u \in (\bfT_i \cup D_i)}
\frac{ h^{(1)}_i ( t; \theta ) }{ \pi_i (s) + h_i ( t ; \theta) }
\end{equation}
Equation~\eqref{eq:WPest} is the estimator suggested
by~\cite{Waagepetersen2008}.  This estimator depends on the
superposition of the event and subsampling processes.
Proposition~\ref{prop:optimal} shows the estimator for~$\theta$
associated with using~\eqref{eq:WPest} is the most efficient within a
suitable class of estimators for the derivative of the cumulative
hazard function (including the Horvitz-Thompson estimator).
Therefore, we restrict our attention to~\eqref{eq:WPest} for the
remainder of this paper. Letting
\begin{equation}
\label{eq:waage_weights}
w_i (t; \theta) = \frac{\pi_i (t)}{\pi_i (t) + h_i (t ; \theta)},
\end{equation}
the resulting estimating equations can be written as
\begin{equation}
\label{eq:approxscore}
\hat{U}_n (\theta) = \sum_{i=1}^n \left[ \sum_{u \in \bfT_i} w_i(u; \theta)
  \frac{h^{(1)} (u; \theta)}{ h ( t; \theta)}  - \sum_{u \in D_i} w_i(u;
  \theta) \frac{h_i^{(1)} (t; \theta)}{ \pi_i (t) } \right].
\end{equation}
Equation~\eqref{eq:approxscore} are approximate score functions built
via plug-in of the design-unbiased estimator of the derivative of the
cumulative hazard given in~\eqref{eq:WPest}.

@@

We also require several assumptions due to the truncation of the
Karhunen-Lo{\`e}ve decomposition that represents~$X(t,s)$.
\begin{assumption}[Functional assumptions~\citep{Park2018}] \normalfont
\label{assumption:truncation}
The following assumptions are standard in prior work on longitudinal
functional data analysis~\citep{Park2018, Yao2005, ChenMuller2012}:
\begin{enumerate}[label=(A.\arabic*)]
\item\label{A1} $X = \{ X(s, T) : (s,T) \in \mathcal{S} \times
  \mathcal{T} \}$ is a square integrable element of the $L^2 (
  \mathcal{S} \times \mathcal{T})$
\item\label{A2} The sampling density~$g(T)$ is continuous
  and~$\sup |g(T)| < \infty$
\item\label{A3} $\E[X(s,T) X(s^\prime, T) X(s,T^\prime)
  X(s^\prime, T^\prime) ] < \infty$ for
  each $s,s^\prime [0,\Delta]$ and~$0 < T, T^\prime < \tau$.
\item\label{A4} $\E[\|X(\cdot, T)\|^4] < \infty$ for each~$0< T
  < \tau$.
\end{enumerate}
\end{assumption}
Assumption~\ref{A3} is guaranteed for by~\ref{E1} and~\ref{E3}, but is
stated here to emphasize its importance for the marginal covariance
procedure of~\cite{Park2018}.
Finally, for simplicity, we assume that there exists~$b^\star$
such that~$\beta(t) = \mathbold{\phi} (t) b^\star$; that is, the
true function~$\beta(t)$ sits in the span of the spline basis
expansion. 
With the necessary assumptions stated, we can now state
Lemma~\ref{lemma:simpleasym}, which provides asymptotic theory.
\begin{lemma} \normalfont
\label{lemma:simpleasym}
Under Assumption~\ref{assumption:events},
Assumption~\ref{assumption:truncation}, and~$\Delta$ known, for
large~$n$ the estimator~$\hat \theta_n$ is consistent; moreover,
\[
\sqrt{n} (\hat \theta - \theta) \to N(0, \Xi (\theta)^{-1})
\]
where
\[
  \Xi (\theta) = \int_{0}^{\tau} w(s; \theta) \left[ \frac{h^{(1)}(s;
      \theta) \times  h^{(1)} (s;\theta)^{\top}}{h(s; \theta)} \right]
  ds.
\]
and~$\tau$ is the random censoring time of the event process.
\end{lemma}
An estimator for~$\Xi(\theta)$ is
\begin{equation}
\label{eq:fisher}
  \hat \Xi (\theta) = n^{-1} \sum_{i=1}^n \sum_{t \in \bfT_i \cup D_i} 
  w_i(t;\theta) (1-w_i(t;\theta)) \left[ \frac{h_i^{(1)}(t;
      \theta)}{h_i (t; \theta)} \right] \times  \left [
    \frac{h_i^{(1)} (t;\theta)}{h_i(t; \theta)} \right]^\top
\end{equation}
For the log-linear intensity model, the sampling-unbiased estimator
for~$\hat \Xi(\theta)$ is equivalent to the Fisher information for the
previously described logistic regression model.
This implies that subsampling from an inhomogeneous Poisson process,
standard logistic regression software can be used to fit the recurrent
event model by specifying an offset equal to~$\log \pi_i (t)$.
Not only this, Lemma~\ref{prop:optimal} shows
weights~\eqref{eq:waage_weights} are optimal within a particular class
of weighted estimating equations.

@@

\subsection{Missing data}
\label{section:missingdata}

@@

Missing data is a critical issue because~$c_{i,k} (t)$ cannot be
estimated if~$X(s,t)$ is not observed for all~$s \in [0,\Delta]$. 
Moreover, we want standard errors to reflect the uncertainty
in these coefficients when missing data is prevalent.
\cite{Goldsmith2011} suggest using best linear unbiased predictors
(BLUP) or posterior modes in the mixed effects model to
estimate~$c_{i,k} (t)$; however, one criticism would be when there is
substantial variability in these estimates.  To deal with this,
\cite{Crainiceanu2010, Goldsmith2010} take a full Bayesian analysis.
\cite{Yao2005} introduced PACE as an alternative frequentist method.
However,~\cite{Petrovich2018} shows that for sparse, irregular
longitudinal, the imputation model should not ignore the outcome
variable~$Y_i (t)$.

Here we present an extension of~\cite{Petrovich2018} to our setting by
leveraging Lemma~\ref{lemma:logistic} and the marginal covariance
estimation procedure to construct a multiple imputation procedure.
Let~$\bfx_{i} (t)$ denote incomplete sensor data at time~$t$ (i.e., at
times~$\{ s_{i,r} \}_{r=1}^{k_{it}} \}$ in~$[0,\Delta]$. Then
\begin{align}
\label{eq:impute}
\E [ X_i (s,t) \given Y_i(t) = y, \bfx_i (t) ] 
  &= \mu_y (s,t) + {\bf a}_{i,t}^\top (s) {\bf B}_{i,t} (\bfx_i (t) -
    \mathbold{\mu}_i (t) ) \\
\text{Cov} \left( X_i (s,t), X_i (s^\prime, t) \given Y_i (t) = y,
  \bfx_i (t) \right) 
  &= \Sigma_{t} (s, s^\prime) - 
    {\bf a}_{i,t} (s)^\top {\bf B}_{i,t} {\bf a}_{i,t} (s^\prime)
\end{align}
where we have
\[
{\bf a}_{i,t} (s)^\top = \left( \begin{array}{c} \Sigma_t (s_{i,1}, s) \\ \vdots \\
                       \Sigma_t (s_{i,k_{it}}, s) \end{array} \right); \quad
{\bf B}_{i,t}^{-1} = \left( 
  \begin{array}{c c c} 
    \Sigma_t(s_{i,1}, s_{i,1}) & \Sigma_t(s_{i,1}, s_{i,2}) & \cdots \\
    \Sigma_t(s_{i,1}, s_{i,2}) & \ddots & \vdots \\
    \vdots & \cdots & \Sigma_t (s_{i,k_{it}}, s_{i,k_{it}}) 
  \end{array} \right),
\]
$\mathbold{\mu}_i (t) = \E [ \bfx_i (t) \given Y_i (t) = y] = \{
\mu_y (s_{j}, t) \}_{j=1}^r$ and~$\mu_y (s,t)$ is the mean of~$X(t,s)$
from group~$y$, and~$\Sigma_t ( s^\prime, s)$ is the covariance
between~$X(s^\prime,t)$ and $X(s,t)$ for $s^\prime, s \in [0, \Delta]$ 
and~$t \in \mathbb{R}_+$.
These covariances can be estimated using the methods considered by
\cite{ChenMuller2012}.  Our imputation procedure instead uses the
marginal covariance~$\Xi$ as a proxy for the true covariance
function~$\Sigma_t(\cdot, \cdot)$.  We found this to work well in
simulations as~$\Sigma_t$ was relatively constant across~$t$; however,
this imputation strategy may be inappropriate when the covariance
structure changes significantly with time~$t$.

@@

@@

We next assess the proposed methodology via a simulation study.
Simulation studies of joint high frequency sensor and recurrent
event data can be quite difficult to construct.  Here, we use the
imputation strategy introduced in Section~\ref{section:missingdata} to
circumvent some of the issues.  
In this section, we assume each individual is observed over five
12-hour days.

First, we generate a sequence of event times via a marginal hazard
rate that ignores sensor data.

We generate 5 days of 12 hours with range of marginal rates (1 per
12 hours, 1 per 6 hours, 1 per hour).  We then sample non-events at
rates or (1 per 12/6/1/0.5 hours).
We assume that~$Y_t = 0$ implies the curve is flat with mean 0.
We assume that~$Y_t = 1$ implies three curves (A) the curve jumps
around 10 minutes prior to the event and remains elevated, (B) the
curve goes down then up and down smoothly, (C) 
These curves are given in Figure~\ref{} for three choices of
parameters that indicate varying degrees of departure from the~$Y_t =
0$ case.

@@

These curves are evaluate

>>>>

An estimator is design-unbiased if its expectation is equal to that
parameter under the probability distribution induced by the sampling
design~\citep{Cassel1977}. Let~$D_i \subset [0,t_i]$ denote the set of
subsampled points.  Note, by design, this set is distinct from the set
of event times, i.e.,~$\bfT_i \cap D_i = \emptyset$.  Under
subsampling via $\pi_i (t)$, one may compute a Horvitz-Thompson
estimator of the derivative of the cumulative hazard:
\[
\hat H_{i}^{(1)} (\tau_i; \theta) = \sum_{u \in D_i} \frac{
  h^{(1)}_i (u; \theta) }{ \pi_i (u) }
\]
An alternative design-unbiased estimator of the derivative of the
cumulative hazards is given by
\begin{equation}
\label{eq:WPest}
\hat H_{i}^{(1)} (\tau_i; \theta) = \sum_{u \in (\bfT_i \cup D_i)}
\frac{ h^{(1)}_i ( u; \theta ) }{ \pi_i (u) + h_i ( u ; \theta) }
\end{equation}
Equation~\eqref{eq:WPest} is the estimator suggested
by~\cite{Waagepetersen2008}.  This estimator depends on the
superposition of the event and subsampling processes.
Proposition~\ref{prop:optimal} shows the estimator for~$\theta$
associated with using~\eqref{eq:WPest} is the most efficient within a
suitable class of estimators for the derivative of the cumulative
hazard function (including the Horvitz-Thompson estimator).
Therefore, we restrict our attention to~\eqref{eq:WPest} for the
remainder of this paper. Letting
\begin{equation}
\label{eq:waage_weights}
w_i (t; \theta) = \frac{\pi_i (t)}{\pi_i (t) + h_i (t ; \theta)},
\end{equation}
the resulting estimating equations can be written as
\begin{equation}
\label{eq:approxscore}
\hat{U}_n (\theta) = \sum_{i=1}^n \left[ \sum_{u \in \bfT_i} w_i(u; \theta)
  \frac{h^{(1)} (u; \theta)}{ h ( u; \theta)}  - \sum_{u \in D_i} w_i(u;
  \theta) \frac{h_i^{(1)} (u; \theta)}{ \pi_i (u) } \right].
\end{equation}
Equation~\eqref{eq:approxscore} are approximate score functions built
via plug-in of the design-unbiased estimator of the derivative of the
cumulative hazard given in~\eqref{eq:WPest}.

@@

We also require several assumptions due to the truncation of the
Karhunen-Lo{\`e}ve decomposition that represents~$X(t,s)$.
\begin{assumption}[Functional assumptions~\citep{Park2018}] \normalfont
\label{assumption:truncation}
The following assumptions are standard in prior work on longitudinal
functional data analysis~\citep{Park2018, Yao2005, ChenMuller2012}:
\begin{enumerate}[label=(A.\arabic*)]
\item\label{A1} $X = \{ X(s, T) : (s,T) \in \mathcal{S} \times
  \mathcal{T} \}$ is a square integrable element of the $L^2 (
  \mathcal{S} \times \mathcal{T})$
\item\label{A2} The sampling density~$g(T)$ is continuous
  and~$\sup |g(T)| < \infty$
\item\label{A3} $\E[X(s,T) X(s^\prime, T) X(s,T^\prime)
  X(s^\prime, T^\prime) ] < \infty$ for
  each $s,s^\prime \in [0,\Delta]$ and~$0 < T, T^\prime < \tau$.
\item\label{A4} $\E[\|X(\cdot, T)\|^4] < \infty$ for each~$0< T
  < \tau$.
\end{enumerate}
\end{assumption}
Assumption~\ref{A3} is guaranteed for by~\ref{E1} and~\ref{E3}, but is
stated here to emphasize its importance for the marginal covariance
procedure of~\cite{Park2018}.
Finally, for simplicity, we assume that there exists~$b^\star$
such that~$\beta(t) = \mathbold{\phi} (t) b^\star$; that is, the
true function~$\beta(t)$ sits in the span of the spline basis
expansion. 
With the necessary assumptions stated, we can now state
Lemma~\ref{lemma:simpleasym}, which provides asymptotic theory.
\begin{lemma} \normalfont
\label{lemma:simpleasym}
Under Assumption~\ref{assumption:events},
Assumption~\ref{assumption:truncation}, and~$\Delta$ known, for
large~$n$ the estimator~$\hat \theta_n$ is consistent; moreover,
\[
\sqrt{n} (\hat \theta - \theta) \to N(0, \Xi (\theta)^{-1})
\]
where
\[
  \Xi (\theta) = \int_{0}^{\tau} w(s; \theta) \left[ \frac{h^{(1)}(s;
      \theta) \times  h^{(1)} (s;\theta)^{\top}}{h(s; \theta)} \right]
  ds.
\]
and~$\tau$ is the random censoring time of the event process.
\end{lemma}
An estimator for~$\Xi(\theta)$ is
\begin{equation}
\label{eq:fisher}
  \hat \Xi (\theta) = n^{-1} \sum_{i=1}^n \sum_{t \in \bfT_i \cup D_i} 
  w_i(t;\theta) (1-w_i(t;\theta)) \left[ \frac{h_i^{(1)}(t;
      \theta)}{h_i (t; \theta)} \right] \times  \left [
    \frac{h_i^{(1)} (t;\theta)}{h_i(t; \theta)} \right]^\top
\end{equation}
For the log-linear intensity model, the sampling-unbiased estimator
for~$\hat \Xi(\theta)$ is equivalent to the Fisher information for the
previously described logistic regression model.
This implies that subsampling from an inhomogeneous Poisson process,
standard logistic regression software can be used to fit the recurrent
event model by specifying an offset equal to~$\log \pi_i (t)$.
Not only this, Lemma~\ref{prop:optimal} shows
weights~\eqref{eq:waage_weights} are optimal within a particular class
of weighted estimating equations.

@@

\subsection{Missing data}
\label{section:missingdata}
\walt{Question: How to perform multiple imputation for~$X(\cdot,t)$
  and $X(\cdot, t^\prime)$ for~$|t - t^\prime| <\Delta$? That is, how
  do you perform multiple imputation with this overlapping structure?}

@@

Missing data is a critical issue because~$c_{i,k} (t)$ cannot be
estimated if~$X(s,t)$ is not observed for all~$s \in [0,\Delta]$. 
Moreover, we want standard errors to reflect the uncertainty
in these coefficients when missing data is prevalent.
\cite{Goldsmith2011} suggest using best linear unbiased predictors
(BLUP) or posterior modes in the mixed effects model to
estimate~$c_{i,k} (t)$; however, one criticism would be when there is
substantial variability in these estimates.  To deal with this,
\cite{Crainiceanu2010} take a full Bayesian analysis.
\cite{Yao2005} introduced PACE as an alternative frequentist method.
However,~\cite{Petrovich2018} shows that for sparse, irregular
longitudinal, the imputation model should not ignore the outcome
variable~$Y_i (t)$.

Here we present an extension of~\cite{Petrovich2018} to our setting by
leveraging Lemma~\ref{lemma:logistic} and the marginal covariance
estimation procedure to construct a multiple imputation procedure.
Let~$\bfx_{i} (t)$ denote incomplete sensor data at time~$t$ (i.e., at
times~$\{ s_{i,r} \}_{r=1}^{k_{it}} \}$ in~$[0,\Delta]$. Then
\begin{align}
\label{eq:impute}
\E [ X_i (s,t) \given Y_i(t) = y, \bfx_i (t) ] 
  &= \mu_y (s,t) + {\bf a}_{i,t}^\top (s) {\bf B}_{i,t} (\bfx_i (t) -
    \mathbold{\mu}_i (t) ) \\
\text{Cov} \left( X_i (s,t), X_i (s^\prime, t) \given Y_i (t) = y,
  \bfx_i (t) \right) 
  &= \Sigma_{t} (s, s^\prime) - 
    {\bf a}_{i,t} (s)^\top {\bf B}_{i,t} {\bf a}_{i,t} (s^\prime)
\end{align}
where we have
\[
{\bf a}_{i,t} (s)^\top = \left( \begin{array}{c} \Sigma_t (s_{i,1}, s) \\ \vdots \\
                       \Sigma_t (s_{i,k_{it}}, s) \end{array} \right); \quad
{\bf B}_{i,t}^{-1} = \left( 
  \begin{array}{c c c} 
    \Sigma_t(s_{i,1}, s_{i,1}) & \Sigma_t(s_{i,1}, s_{i,2}) & \cdots \\
    \Sigma_t(s_{i,1}, s_{i,2}) & \ddots & \vdots \\
    \vdots & \cdots & \Sigma_t (s_{i,k_{it}}, s_{i,k_{it}}) 
  \end{array} \right),
\]
$\mathbold{\mu}_i (t) = \E [ \bfx_i (t) \given Y_i (t) = y] = \{
\mu_y (s_{j}, t) \}_{j=1}^r$ and~$\mu_y (s,t)$ is the mean of~$X(t,s)$
from group~$y$, and~$\Sigma_t ( s^\prime, s)$ is the covariance
between~$X(s^\prime,t)$ and $X(s,t)$ for $s^\prime, s \in [0, \Delta]$ 
and~$t \in \mathbb{R}_+$.
These covariances can be estimated using the methods considered by
\cite{ChenMuller2012}.  
% Our imputation procedure instead uses the
% marginal covariance~$\Xi$ as a proxy for the true covariance
% function~$\Sigma_t(\cdot, \cdot)$.  
% We found this to work well in simulations as~$\Sigma_t$ was relatively
% constant across~$t$; however, this imputation strategy may be
% inappropriate when the covariance structure changes significantly with
% time~$t$.

@@

\newpage

@@

We next assess the proposed methodology via a simulation study.
Simulation studies of joint high frequency sensor and recurrent
event data can be quite difficult to construct.  Here, we use the
imputation strategy introduced in Section~\ref{section:missingdata} to
circumvent some of these issues.

Here, we assume each individual is observed over five 12-hour days.  
We then generate a sequence of event times via a marginal hazard
rate that ignores sensor data.  We assume the marginal rate is
constant and set to 1 per 12/6/1 hour(s). We then sample non-events at
rates of 1 per 12/6/1/0.5 hours.
This sampling strategy yields (non-)event times~$t$ along with
labels~$Y_t \in \{0,1\}$. 
We then use the imputation strategy given by
equation~\eqref{eq:impute} to generate the curves~$\{ X(s,t) \}$.
For simplicity, we do not deal with 
We assume that~$Y_t = 0$ implies the curve is flat with mean 0.
We assume that~$Y_t = 1$ implies three curves (A) the curve jumps
around 10 minutes prior to the event and remains elevated, (B) the
curve goes down then up and down smoothly, (C) 
These curves are given in Figure~\ref{} for three choices of
parameters that indicate varying degrees of departure from the~$Y_t =
0$ case.
The associated~$\beta$ depends on the covariance~$\Sigma_t (s,t)$.
In particular, $\beta = \Sigma_t^{-1} (\mu_{1,t} - \mu_{0,t})$ is
guaranteed by Lemma~\ref{} and Theorem 1 of \cite{Petrovich2018}.
For~$\beta(s)$ non time-varying implies the covariance term

@@

to estimate the impact of strong X dependence for close X's then we
will GENERATE SEQUENTIALLY TO FILL IN TO ACCOUNT FOR OVERLAP!
HOW SHOULD WE HANDLE COVARIANCE BETWEEN X's outside the range?
WE WILL APPROXIMATE?
